Deciding Bitvector Termination with Program Synthesis

### 2014-07-04 papers/termination/paper.tex
Soundness & completeness

\begin{theorem}
 {\sc Headshot} is complete -- if a program terminates, {\sc Headshot} will find a proof that it does.
\end{theorem}

>>>>

\begin{proof}
 The ranking function constitutes a constructive, machine checked proof that the program terminates.
\end{proof}

\begin{theorem}
 {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
\end{theorem}

\begin{proof}
 If the program under analysis $P$ has variables $v_1, \ldots, v_k$ each of which is $b$ bits wide, its state space $\mathcal{S}$ is of size $2^{bk}$.
 A ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$ exists iff $P$ terminates.  Since $R$ is injective, we have that
 $\| \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \| \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to \mathcal{D'}$
 with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting $R' = R|_\mathcal{S}$, i.e. $R'$ is just the restriction of $R$ to $\mathcal{S}$
 Since $\mathcal{S}$ already comes equipped with a natural well ordering we can also construct $R'' = \iota \circ R'$
 where $\iota: \mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from $\mathcal{D'}$ to $\mathcal{S}$.
 So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.

 It is fairly easy to see that any finite permutation is computed by a finite program in the proof language generated
 by {\sc Kalashnikov}.  A very inefficient construction which computes the permutation $R(x) = x+1$ for one variable
 is:

 \begin{verbatim}
t1 = 0
t2 = v1 == 0
t3 = ITE(t2, 1, t1)
t4 = v1 == 1
t5 = ITE(t4, 2, t3)
...
 \end{verbatim}

If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$.

So we can see that every permutation is computed a program of size at most $b \times 2^{bk + 1}$.  Since
{\sc Kalashnikov} enumerates all programs in order of length, it will eventually check
a program computing each permutation of $\mathcal{S}$ and so will eventually check a program
computing the ranking function $R''$.  Checking that a single program satisfies the specification
is a decidable problem, and the search as a whole terminates.
\end{proof}
